---
title: "Assignment 2: Coding Basics"
author: "Ben Egan"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on coding basics. 

## Directions
1. Change "Student Name" on line 3 (above) with your name.
2. Work through the steps, **creating code and output** that fulfill each instruction.
3. Be sure to **answer the questions** in this assignment document.
4. When you have completed the assignment, **Knit** the text and code into a single PDF file.
5. After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai. Add your first and last name into the file name (e.g., "FirstLast_A02_CodingBasics.Rmd") prior to submission.



## Basics Day 1

1. Generate a sequence of numbers from one to 100, increasing by fours. Assign this sequence a name. 

2. Compute the mean and median of this sequence. 

3. Ask R to determine whether the mean is greater than the median. 

4. Insert comments in your code to describe what you are doing.

```{r}
#1. 
sequence4 <- function(x){seq(1,100,4)}

#2. 
mean(sequence4())
median(sequence4())


#3. 
ifelse(mean(sequence4())>median(sequence4()),true,FALSE)
#I have asked r to tell me simply if the mean of sequence4 is greater than the median of sequence4.
```

## Basics Day 2

5. Create a series of vectors, each with four components, consisting of (a) names of students, (b) test scores out of a total 100 points, and (c) whether or not they have passed the test (TRUE or FALSE) with a passing grade of 50.

6. Label each vector with a comment on what type of vector it is.

7. Combine each of the vectors into a data frame. Assign the data frame an informative name.

8. Label the columns of your data frame with informative titles.

```{r}
Names <- c("john","adam","sarah","Emily")
#created a list of random names

scores <- c(95,82,75,60)
#created a list of random scores

Passed <- c("true", "false", "false", "true")
#assigned passing grade based on above list

create_df <- data.frame("names"=Names,"scores"= scores, "passed"=Passed)
create_df
#created data frame

```

9. QUESTION: How is this data frame different from a matrix?

> Answer: A data frame is different from a matrix as it contains numerical values as well as character values. A matrix just contains values of all the same data type

10. Create a function with an if/else statement. Your function should determine whether a test score is a passing grade of 50 or above (TRUE or FALSE). You will need to choose either the `if` and `else` statements or the `ifelse` statement. Hint: Use `print`, not `return`. The name of your function should be informative.

11. Apply your function to the vector with test scores that you created in number 5. 

```{r}
Cond_names <- ifelse(scores>50,"true","false")
#created an ifelse statement to determine if a score is passing or not

Test_Results <- data.frame("names"=Names,"scores"= scores, "passed"= Cond_names)
#made a new Data Frame with the new function in place of the old list of whether or not the student passed

Test_Results



```

12. QUESTION: Which option of `if` and `else` vs. `ifelse` worked? Why?

> Answer: 
> print(if(scores > 50) {TRUE} else {FALSE})
#Ifelse worked better as it could handle more than one argument. The if...else setup only returned the virst row in the Dataframe.

