---
title: 'Assignment 3: Data Exploration'
author: "Ben Egan, Section Thursday"
output:
  pdf_document: default
  word_document: default
  html_document:
    df_print: paged
geometry: margin=2.54cm
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Exploration. 

## Directions
1. Change "Student Name, Section #" on line 3 (above) with your name and section number.

2. Work through the steps, **creating code and output** that fulfill each instruction.

3. Be sure to **answer the questions** in this assignment document.

4. When you have completed the assignment, **Knit** the text and code into a single PDF file.
5. After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai. Add your last name into the file name (e.g., "FirstLast_A03_DataExploration.Rmd") prior to submission.

The completed exercise is due on <1.31>.


## Set up your R session

1. Check your working directory, load necessary packages (tidyverse), and upload two datasets: the ECOTOX neonicotinoid dataset (ECOTOX_Neonicotinoids_Insects_raw.csv) and the Niwot Ridge NEON dataset for litter and woody debris (NEON_NIWO_Litter_massdata_2018-08_raw.csv). Name these datasets "Neonics" and "Litter", respectively. **Be sure to add the `stringsAsFactors = TRUE` parameter to the function when reading in the CSV files.**
```{r,  message = FALSE}
Neonics <- read.csv(file.choose(), stringsAsFactors = TRUE)
Litter <- read.csv(file.choose(), stringsAsFactors = TRUE)
library(ggplot2)
library(tidyverse)

```

## Learn about your system

2. The neonicotinoid dataset was collected from the Environmental Protection Agency's ECOTOX Knowledgebase, a database for ecotoxicology research. Neonicotinoids are a class of insecticides used widely in agriculture. The dataset that has been pulled includes all studies published on insects. Why might we be interested in the ecotoxicologoy of neonicotinoids on insects? Feel free to do a brief internet search if you feel you need more background information. 

> Answer: Neonicotinoids are an effective inseticide and unfortunately that effect is fealt by Bee populations as well. Ecotoxicolgy data is important to see how the use of these insectides may be impacting Bee populations, water health and other envrionmental implications

3. The Niwot Ridge litter and woody debris dataset was collected from the National Ecological Observatory Network, which collectively includes 81 aquatic and terrestrial sites across 20 ecoclimatic domains. 32 of these sites sample forest litter and woody debris, and we will focus on the Niwot Ridge long-term ecological research (LTER) station in Colorado. Why might we be interested in studying litter and woody debris that falls to the ground in forests? Feel free to do a brief internet search if you feel you need more background information. 

> Answer: Litter and woody debris are great sources of food for organisms/ bacteria/ mosses on the forest floor. This in turn creates healthy soil and healthy plants. Litter and woody debris can also drastically reduce soil erosion.

4. How is litter and woody debris sampled as part of the NEON network? Read the NEON_Litterfall_UserGuide.pdf document to learn more. List three pieces of salient information about the sampling methods here: 

> Answer: 
*The sampling of the Litter and Woody Debris happens in tower plots where debris falls into a marked off area that can be easily measured
*Some of the traps are elevated off the ground and some are on the ground)
*Depending on the the trap method and how wooded the are is, the sites can be sampled every 2 weeks to  6 Months



## Obtain basic summaries of your data (Neonics)

5. What are the dimensions of the dataset?

```{r}
dim(Neonics)
```

6. Using the `summary` function on the "Effect" column, determine the most common effects that are studied. Why might these effects specifically be of interest?
```{r, include=TRUE}
summary(Neonics$Effect)
```

> Answer: Some of the most common are Mortality and Popultaion. It's important to know this as the scientist are trying to measure the effect of the pesticides on eliminating Pests

7. Using the `summary` function, determine the six most commonly studied species in the dataset (common name). What do these species have in common, and why might they be of interest over other insects? Feel free to do a brief internet search for more information if needed.

```{r}
summary(Neonics$Species.Common.Name)
```

> Answer: The six most common Are Honey Bee, Parasitic Wasp, Buff Tailed Bumblebee, Carniolan Honey Bee, and Bumble Bee. They are all Bees which are crucial to the pollination of the crops and the general health of the ecological environment.
                               

8. Concentrations are always a numeric value. What is the class of Conc.1..Author. in the dataset, and why is it not numeric?

```{r}
class(Neonics$Conc.1..Author)
```

> Answer: The class is "factor" because some of the values have been entered as NR or with "/". If there is a non-numeric value than the whole row is recorded as Character



## Explore your data graphically (Neonics)

9. Using `geom_freqpoly`, generate a plot of the number of studies conducted by publication year. 
```{r,  message = FALSE}
ggplot(Neonics) +
  geom_freqpoly(aes(x = Publication.Year), bins = 50)

```

10. Reproduce the same graph but now add a color aesthetic so that different Test.Location are displayed as different colors.
```{r,  message = FALSE}
#ggplot(Neonics, aes(x = Publication.Year, color = Test.Location)) +
  #geom_freqpoly() + theme(legend.position = "top")
  
```

Interpret this graph. What are the most common test locations, and do they differ over time?

> Answer: The most common are lab and Field Natural. And over time, Lab tests become much more popular while natural studies have come down a little since the mid 00's.

11. Create a bar graph of Endpoint counts. What are the two most common end points, and how are they defined? Consult the ECOTOX_CodeAppendix for more information. 

```{r,  message = FALSE}
ggplot(Neonics, aes(x = Endpoint), color = Endpoint) + geom_bar()+theme(axis.text.x = element_text(angle =  90, vjust = .5, hjust = 1))

```

> Answer: 
>The two Most common endpoints are LOEL And NOEL. a LOEL is the lowest observed effect level and the NOEL is the No Observed effect Level. For Toxicology studies these are teh most important distinguishers as they determine the to what the standards.

## Explore your data (Litter)

12. Determine the class of collectDate. Is it a date? If not, change to a date and confirm the new class of the variable. Using the `unique` function, determine which dates litter was sampled in August 2018. 
```{r,  message = FALSE} 
class(Litter$collectDate)

Litter$collectDate <- as.Date(Litter$collectDate, format = "%Y-%m-%d")

Aug_Dates <- unique(Litter$collectDate)

print(Aug_Dates)

```

13. Using the `unique` function, determine how many plots were sampled at Niwot Ridge. How is the information obtained from `unique` different from that obtained from `summary`?

```{r,  message = FALSE}
Niwot_Ridge <- unique(Litter$plotID)
print(Niwot_Ridge)

sum(summary(Niwot_Ridge))
```

> Answer: They are different as the Unique function lists all of the individual Plot names. My summary function has summed the unique values and leaves out the individual names.

14. Create a bar graph of functionalGroup counts. This shows you what type of litter is collected at the Niwot Ridge sites. Notice that litter types are fairly equally distributed across the Niwot Ridge sites.
```{r,  message = FALSE}
ggplot(Litter, aes(x = functionalGroup), color = functionalGroup) + geom_bar()
```

15. Using `geom_boxplot` and `geom_violin`, create a boxplot and a violin plot of dryMass by functionalGroup. 
```{r,  message = FALSE}
ggplot(Litter) + geom_boxplot(aes(x = functionalGroup, y = dryMass))

ggplot(Litter) + geom_boxplot(aes(x = functionalGroup, y = dryMass))

ggplot(Litter) +
  geom_violin(aes(x = functionalGroup, y = dryMass), 
              draw_quantiles = c(0.25, 0.5, 0.75))
```

Why is the boxplot a more effective visualization option than the violin plot in this case?

> Answer: The Box plot gives a lot more information that shows outliers and the quartiles. the Violin does not give the quartile information and distributes the full range evenly

What type(s) of litter tend to have the highest biomass at these sites? 

> Answer: Needles have the highest biomass at these sites.